---
title: "Space-Time Prediction of Bike Share Demand"
author: "Palak Agarwal"
date: "November 13, 2020"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: "hide"
    code_download: true
---

## 1 Introduction

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The San Francisco Municipal Transportation Agency (SFMTA) is a department of the City and County of San Francisco responsible for the management of all ground transportation in the city. The SFMTA has oversight over the Municipal Railway (Muni) public transit, as well as bicycling, paratransit, parking, traffic, walking, and taxis. In 2013 they partnered with public agencies to plan and implement the original Bay Area Bike Share pilot project and are now leading San Francisco’s efforts to work with the private sector partner to expand the system to over ten times its pilot size. They patnered with Lyft to expand the regional system which has since been re-branded with all-new equipment as Bay Wheels. In 2017, as independent stationless bikeshare emerged as a big new trend, they were among the first U.S. cities to create a regulatory and permitting framework to address this fast-moving phenomenon and insure that bikeshare in all its forms is safe, orderly and equitable for all San Franciscans.

One of the big operational challenges of bike share systems is "re-balancing" - getting bikes to stations that are anticipated to have demand but lack bikes. Figuring out how to do this is one of the keys to operating a successful system. With the introduction of the stationless bikeshare system the problem of re-balancing can be eliminated by a certain percentage. But it creates new problems as the bikes are further spread out and its harder to make sure that the bike is picked up. For the assignment, the data used is only from docked stations and does not include the stationless bikes.

For making the bike-share system more efficient i.e. to re-balance it many strategies can be introduced. I think the most effect strategy would be to use a reward system. If we know the capacity of a bike station and knew that the user was going to that station we could incentivize them to drop off the bike to another bike station or we could also incentivize them to use a bike which has not been touched for more than a two weeks. This will allow for an efficient demand and supply system.

Today's trend is similar to the trend tomorrow and this week's trend will be similar to next week trends. Hence, we will be able to predict how many bikes and from which station to high accuracy as you will see in the later in the model and assignment. If we know how many rides are going to take place, it is easier to use that data to manipulate which bikes are being used and which are not. For example, the Financial District of downtown SF is very busy and has many stations in the same block. If for some reason one station is more active than the other, we can use rewards to make people pick and drop bikes from different stations. 

Lyft using its platform and app to show the users where the bikes are available so that can be used to make users pick up bikes that have not been touched for a while and also have them drop it off to locations where the demand is high and there is a shortage of bikes. SFMTA already has concessions for different individuals and communities which the bike share system can tap into as well. Knowing the capacity of a docking station will help us know if it has any open slots or not which can be a factor to provide users with offers to drop the bike off at other locations or if a busy dock is empty the user can be incentivized to drop off the bike there. 

## 2.0 Setup
### 2.1 Libraries and themes
Let's load relevant libraries and some graphic themes.
```{r setup, cache=TRUE, message=FALSE}
## Setup
library(tidyverse)
library(sf)
library(lubridate)
library(tigris)
library(tidycensus)
library(viridis)
library(riem)
library(gridExtra)
library(knitr)
library(kableExtra)
library(RSocrata)
library(FNN)

mapTheme <- function(base_size = 12) {
  theme(
    text = element_text( color = "black"),
    plot.title = element_text(size = 24,colour = "black"),
    plot.subtitle=element_text(face="italic"),
    plot.caption=element_text(hjust=0),
    axis.ticks = element_blank(),
    panel.background = element_blank(),axis.title = element_blank(),
    axis.text = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill=NA, size=2)
  )
}

plotTheme <- function(base_size = 12) {
  theme(
    text = element_text( color = "black"),
    plot.title = element_text(size = 24,colour = "black"),
    plot.subtitle = element_text(face="italic"),
    plot.caption = element_text(hjust=0),
    axis.ticks = element_blank(),
    panel.background = element_blank(),
    panel.grid.major = element_line("grey60", size = 0.1),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill=NA, size=2),
    strip.background = element_rect(fill = "grey60", color = "white"),
    strip.text = element_text(size=12),
    axis.title = element_text(size=12),
    axis.text = element_text(size=10),
    plot.background = element_blank(),
    legend.background = element_blank(),
    legend.title = element_text(colour = "black", face = "italic"),
    legend.text = element_text(colour = "black", face = "italic"),
    strip.text.x = element_text(size = 14)
  )
}

# Load Quantile break functions

qBr <- function(df, variable, rnd) {
  if (missing(rnd)) {
    as.character(quantile(round(df[[variable]],0),
                          c(.01,.2,.4,.6,.8), na.rm=T))
  } else if (rnd == FALSE | rnd == F) {
    as.character(formatC(quantile(df[[variable]]), digits = 3),
                 c(.01,.2,.4,.6,.8), na.rm=T)
  }
}

q5 <- function(variable) {as.factor(ntile(variable, 5))}

palette5 <- c("#eff3ff","#bdd7e7","#6baed6","#3182bd","#08519c")
palette4 <- c("#D2FBD4","#92BCAB","#527D82","#123F5A")
palette2 <- c("#6baed6","#08519c")

# Functions
nn_function <- function(measureFrom,measureTo,k) {
  measureFrom_Matrix <- as.matrix(measureFrom)
  measureTo_Matrix <- as.matrix(measureTo)
  nn <-   
    get.knnx(measureTo, measureFrom, k)$nn.dist
  output <-
    as.data.frame(nn) %>%
    rownames_to_column(var = "thisPoint") %>%
    gather(points, point_distance, V1:ncol(.)) %>%
    arrange(as.numeric(thisPoint)) %>%
    group_by(thisPoint) %>%
    summarize(pointDistance = mean(point_distance)) %>%
    arrange(as.numeric(thisPoint)) %>% 
    dplyr::select(-thisPoint) %>%
    pull()
  
  return(output)  
}
```

### 2.2 Data Wrangling
Reading in the bike data.
```{r load_data, message = FALSE, warning=FALSE, results=FALSE}
## Read data
sf_bike1 <- read.csv('Data/201803-fordgobike.csv')
sf_bike2 <- read.csv('Data/201804-fordgobike.csv')

sf_bike <- rbind(sf_bike1, sf_bike2)
glimpse(sf_bike)

sfCounty <-
  st_read("https://data.sfgov.org/api/geospatial/p5b7-5n3h?method=export&format=GeoJSON") %>% 
  st_union() %>%
  st_transform('ESRI:102241')

neighborhoods <- 
  st_read("https://data.sfgov.org/api/geospatial/pty2-tcw4?method=export&format=GeoJSON") %>%
  st_transform('ESRI:102241') %>%
  dplyr::select(-link)
```
After that we use the data to create bins of 15 and 60 minute intervals by rounding.
```{r time_bins, cache = TRUE}
sfBike <- sf_bike %>%
  mutate(interval60 = floor_date(ymd_hms(start_time), unit = "hour"),
         interval15 = floor_date(ymd_hms(start_time), unit = "15 mins"),
         week = week(interval60),
         dotw = wday(interval60, label=TRUE)) %>%
  filter(week < 15)
```

### 2.3 Census Data
For the analysis, I am using 2018 census data for San Francisco and Alameda county.
```{r census, message = FALSE, results=FALSE}
# Install Census API Key
census_api_key("aea3dee2d96acb5101e94f3dcfa1b575f73d093a", overwrite = TRUE)

sfCensus <- 
  get_acs(geography = "tract", 
          variables = c("B01003_001", "B19013_001", 
                        "B02001_002", "B08013_001",
                        "B08012_001", "B08301_001", 
                        "B08301_010", "B01002_001"), 
          year = 2018, 
          state = 06, 
          geometry = TRUE, 
          county=c(001,075),
          output = "wide") %>%
  rename(Total_Pop =  B01003_001E,
         Med_Inc = B19013_001E,
         Med_Age = B01002_001E,
         White_Pop = B02001_002E,
         Travel_Time = B08013_001E,
         Num_Commuters = B08012_001E,
         Means_of_Transport = B08301_001E,
         Total_Public_Trans = B08301_010E) %>%
  select(Total_Pop, Med_Inc, White_Pop, Travel_Time,
         Means_of_Transport, Total_Public_Trans,
         Med_Age,
         GEOID, geometry) %>%
  mutate(Percent_White = White_Pop / Total_Pop,
         Mean_Commute_Time = Travel_Time / Total_Public_Trans,
         Percent_Taking_Public_Trans = Total_Public_Trans / Means_of_Transport) %>%
  st_transform('ESRI:102241')

sfTracts <- 
  sfCensus %>%
  as.data.frame() %>%
  distinct(GEOID, .keep_all = TRUE) %>%
  select(GEOID, geometry) %>% 
  st_sf
```

## 2.4 Connect census data to the the Bikeshare data
Now lets add the spatial information to our bikeshare data as origin and destination data, first joining the origin station, then the destination station to our census data.
```{r add_census_tracts, cache = TRUE, message = FALSE, warning = FALSE}
## Adding census data to bike data
sfBike_census <- st_join(sfBike %>% 
                        filter(is.na(start_station_longitude) == FALSE &
                                 is.na(start_station_latitude) == FALSE &
                                 is.na(end_station_latitude) == FALSE  &
                                 is.na(end_station_longitude) == FALSE) %>%
                        st_as_sf(., coords = c("start_station_longitude", "start_station_latitude"), crs = 4326),
                      sfTracts %>%
                        st_transform(crs=4326),
                      join=st_intersects,
                          left = TRUE) %>%
  rename(Origin.Tract = GEOID) %>%
  mutate(start_station_longitude = unlist(map(geometry, 1)),
         start_station_latitude = unlist(map(geometry, 2)))%>%
  as.data.frame() %>%
  select(-geometry)%>%
  st_as_sf(., coords = c("end_station_longitude", "end_station_latitude"), crs = 4326) %>%
  st_join(., sfTracts %>%
            st_transform(crs=4326),
          join=st_intersects,
          left = TRUE) %>%
  rename(Destination.Tract = GEOID)  %>%
  mutate(end_station_longitude = unlist(map(geometry, 1)),
         end_station_latitude = unlist(map(geometry, 2)))%>%
  as.data.frame() %>%
  na.omit(Origin.Tract, Destination.Tract) %>%
  select(-geometry)
```

## 2.5 Weather Data
```{r import_weather, message = FALSE, warning = FALSE, cache = TRUE}
## Weather Data

weather.Panel <- 
  riem_measures(station = "SFO", date_start = "2018-03-01", date_end = "2018-04-30") %>%
  dplyr::select(valid, tmpf, p01i, sknt)%>%
  replace(is.na(.), 0) %>%
  mutate(interval60 = ymd_h(substr(valid,1,13))) %>%
  mutate(week = week(interval60),
         dotw = wday(interval60, label=TRUE)) %>%
  group_by(interval60) %>%
  summarize(Temperature = max(tmpf),
            Precipitation = sum(p01i),
            Wind_Speed = max(sknt)) %>%
  mutate(Temperature = ifelse(Temperature == 0, 42, Temperature))

grid.arrange(
  ggplot(weather.Panel, aes(interval60,Precipitation)) + geom_line() + 
    labs(title="Percipitation", x="Hour", y="Perecipitation") + plotTheme(),
  ggplot(weather.Panel, aes(interval60,Wind_Speed)) + geom_line() + 
    labs(title="Wind Speed", x="Hour", y="Wind Speed") + plotTheme(),
  ggplot(weather.Panel, aes(interval60,Temperature)) + geom_line() + 
    labs(title="Temperature", x="Hour", y="Temperature") + plotTheme(),
  top="Weather Data - San Francisco SFO - March-April, 2018")
```

## 3 Data Exploration

We begin by examining the time and frequency components of our data.

First, we look at the overall time pattern - there is clearly a daily periodicity and there are lull periods on weekends.
```{r plot1, message = FALSE, warning = FALSE, cache = TRUE}
ggplot(sfBike_census %>%
         group_by(interval60) %>%
         tally())+
  geom_line(aes(x = interval60, y = n))+
  labs(title="Bike share trips per hr, SF + Alameda, March, 2018",
       x="Date", 
       y="Number of trips")+
  plotTheme()
```

```{r plot2, message = FALSE, warning = FALSE, cache = TRUE}
## Mean number of hourly trips
sfBike_census %>%
  mutate(time_of_day = case_when(hour(interval60) < 7 | hour(interval60) > 18 ~ "Overnight",
                                 hour(interval60) >= 7 & hour(interval60) < 10 ~ "AM Rush",
                                 hour(interval60) >= 10 & hour(interval60) < 15 ~ "Mid-Day",
                                 hour(interval60) >= 15 & hour(interval60) <= 18 ~ "PM Rush"))%>%
  group_by(interval60, start_station_name, time_of_day) %>%
  tally()%>%
  group_by(start_station_name, time_of_day)%>%
  summarize(mean_trips = mean(n))%>%
  ggplot()+
  geom_histogram(aes(mean_trips), binwidth = 1)+
  labs(title="Mean Number of Hourly Trips Per Station. SF+Alameda, March-April, 2018",
       x="Number of trips", 
       y="Frequency")+
  facet_wrap(~time_of_day)+
  plotTheme()
```

```{r plot3, message = FALSE, warning = FALSE, cache = TRUE}
## By station
ggplot(sfBike_census %>%
         group_by(interval60, start_station_name) %>%
         tally())+
  geom_histogram(aes(n), binwidth = 5)+
  labs(title="Bike share trips per hr by station. SF+Alameda, March-April, 2018",
       x="Number of Stations", 
       y="Trip Counts")+
  plotTheme()
```

```{r plot4, message = FALSE, warning = FALSE, cache = TRUE}
## Number of trips by hr in a week
ggplot(sfBike_census %>% mutate(hour = hour(start_time)))+
  geom_freqpoly(aes(hour, color = dotw), binwidth = 1)+
  labs(title="Bike share trips by day of the week, SF+Alameda, March-April, 2018",
       x="Hour", 
       y="Trip Counts")+
  plotTheme()
```

```{r plot5, message = FALSE, warning = FALSE, cache = TRUE}
## Number of trips weekday - weekend
ggplot(sfBike_census %>% 
         mutate(hour = hour(start_time),
                weekend = ifelse(dotw %in% c("Sun", "Sat"), "Weekend", "Weekday")))+
  geom_freqpoly(aes(hour, color = weekend), binwidth = 1)+
  labs(title="Bike share trips - weekend vs weekday, SF+Alameda, March-April, 2018",
       x="Hour", 
       y="Trip Counts")+
  plotTheme()
```


```{r plot6, message = FALSE, warning = FALSE, cache = TRUE}
## Rush by station on census tracts
ggplot()+
  geom_sf(data = sfTracts %>%
            st_transform(crs=4326), colour = '#efefef')+
  geom_point(data = sfBike_census %>% 
               mutate(hour = hour(start_time),
                      weekend = ifelse(dotw %in% c("Sun", "Sat"), "Weekend", "Weekday"),
                      time_of_day = case_when(hour(interval60) < 7 | hour(interval60) > 18 ~ "Overnight",
                                              hour(interval60) >= 7 & hour(interval60) < 10 ~ "AM Rush",
                                              hour(interval60) >= 10 & hour(interval60) < 15 ~ "Mid-Day",
                                              hour(interval60) >= 15 & hour(interval60) <= 18 ~ "PM Rush"))%>%
               group_by(start_station_id, start_station_latitude, start_station_longitude, weekend, time_of_day) %>%
               tally(),
             aes(x=start_station_longitude, y = start_station_latitude, color = n), 
             fill = "transparent", alpha = 0.4, size = 1.5)+
  scale_colour_viridis(direction = -1,
                       discrete = FALSE, option = "D")+
  ylim(min(sfBike_census$start_station_latitude), max(sfBike_census$start_station_latitude))+
  xlim(min(sfBike_census$start_station_longitude), max(sfBike_census$start_station_longitude))+
  facet_grid(weekend ~ time_of_day)+
  labs(title="Bike share trips per hr by station. Chicago, May, 2018")+
  mapTheme()
```

## 4 Create Space-Time Panel

First **we have to make sure each unique station and hour/day combo exists in our data set.** This is done in order to create a "panel" (e.g. a time-series) data set where each time period in the study is represented by a row - whether an observation took place then or not. So if a station didn't have any trips originating from it at a given hour, we still need a zero in that spot in the panel.

We start by determining the maximum number of combinations.

Then we compare that to the actual number of combinations. We create an empty data frame `study.panel`, is created that has each unique space/time observations. This is done using the expand.grid function and unique. Along the way, we keep tabs on the number of rows our data have - `nrow` shows that the count is still correct.

We then join the station name, tract and lat/lon (some have multiple lat lon info, so we just take the first one of each using `group_by` and `slice`).
We create the full panel by summarizing counts by station for each time interval, keep census info and lat/lon information along for joining later to other data. We remove data for station IDs that are `FALSE`.

We also ditch a bit of data (this is why `study.panel` and `ride.panel` don't end up being exactly the same length). There are two stations - Dusable harbor and Eastlake Terrace that don't join properly to census tracts. They are too close to the water and don't play nice with our tracts. In the service of laziness, we get rid of these.

```{r panel_length_check, cache = TRUE, message = FALSE, warning = FALSE}
length(unique(sfBike_census$interval60)) * length(unique(sfBike_census$start_station_id))

study.panel <- 
  expand.grid(interval60=unique(sfBike_census$interval60), 
              start_station_id = unique(sfBike_census$start_station_id)) %>%
  left_join(., sfBike_census %>%
              select(start_station_id, start_station_name, Origin.Tract, start_station_longitude, start_station_latitude )%>%
              distinct() %>%
              group_by(start_station_id) %>%
              slice(1))

nrow(study.panel)      

## Ride Panel
ride.panel <- 
  sfBike_census %>%
  mutate(Trip_Counter = 1) %>%
  right_join(study.panel) %>% 
  group_by(interval60, start_station_id, start_station_name, Origin.Tract, start_station_longitude, start_station_latitude) %>%
  summarize(Trip_Count = sum(Trip_Counter, na.rm=T)) %>%
  left_join(weather.Panel) %>%
  ungroup() %>%
  filter(is.na(start_station_id) == FALSE) %>%
  mutate(week = week(interval60),
         dotw = wday(interval60, label = TRUE)) %>%
  filter(is.na(Origin.Tract) == FALSE)

ride.panel <- 
  left_join(ride.panel, sfCensus %>%
              as.data.frame() %>%
              select(-geometry), by = c("Origin.Tract" = "GEOID"))
```

## 5 Feature Engineering

### 5.1 Time Lags

Creating time lag variables will add additional nuance about the demand during a given time period - hours before and during that day. 

We can also try to control for the effects of holidays that disrupt the expected demand during a given weekend or weekday. We have a holiday on May 28 - Memorial Day. For that three day weekend we could use some dummy variables indicating temporal proximity to the holiday.

Keep in mind, that unique fixed effects must be in your training set when you run your models.

We can evaluate the correlations in these lags. They are pretty strong. There's a Pearson's R of 0.84 for the `lagHour` - that's very, very strong. 

This makes a lot of intuitive sense - the demand right now should be relatively similar to the demand tomorrow at this time, and to the demand an hour from now, but twelve hours from now, we likely expect the opposite in terms of demand.

```{r time_lags, cache = TRUE, message = FALSE}
ride.panel <- 
  ride.panel %>% 
  arrange(start_station_id, interval60) %>% 
  mutate(lagHour = dplyr::lag(Trip_Count,1),
         lag2Hours = dplyr::lag(Trip_Count,2),
         lag3Hours = dplyr::lag(Trip_Count,3),
         lag4Hours = dplyr::lag(Trip_Count,4),
         lag12Hours = dplyr::lag(Trip_Count,12),
         lag1day = dplyr::lag(Trip_Count,24)) %>%
  mutate(day = yday(interval60))

ride.panel <-
  ride.panel %>%
  mutate(X = start_station_longitude, Y = start_station_latitude )%>%
  st_as_sf(coords = c("X", "Y"), crs = 4326, agr = "constant") %>%
  st_transform('ESRI:102241')
```

```{r evaluate_lags, cache = TRUE, warning = FALSE, message = FALSE}
as.data.frame(ride.panel) %>%
  group_by(interval60) %>% 
  summarise_at(vars(starts_with("lag"), "Trip_Count"), mean, na.rm = TRUE) %>%
  gather(Variable, Value, -interval60, -Trip_Count) %>%
  mutate(Variable = factor(Variable, levels=c("lagHour","lag2Hours","lag3Hours","lag4Hours",
                                              "lag12Hours","lag1day")))%>%
  group_by(Variable) %>%  
  summarize(correlation = round(cor(Value, Trip_Count),2))
```


### 5.2 Exposure Features
```{r exposure, cache = TRUE, message = FALSE}
##Exposure Features 

### Parks
parks_al <- 
  read.socrata("https://data.oaklandca.gov/resource/kq8i-6bzk.json") %>%
  dplyr::select(Y = location_1.latitude, X = location_1.longitude) %>%
  na.omit() %>%
  st_as_sf(coords = c("X", "Y"), crs = 4326, agr = "constant") %>%
  st_transform('ESRI:102241') %>%
  mutate(Legend = "Park")

parks_sf <-
  st_read("C:/Users/agarw/Documents/MUSA508/MUSA508-Assignment3/Data/Recreation_and_Parks_Properties.csv") %>%
  dplyr::select(Y = Latitude, X = Longitude) %>%
  na.omit() %>%
  st_as_sf(coords = c("X", "Y"), crs = 4326, agr = "constant") %>%
  st_transform('ESRI:102241') %>%
  mutate(Legend = "Park")

parks <- rbind(parks_al, parks_sf)

ride.panel <-
  ride.panel %>% 
  mutate(
    park_dist = nn_function(st_coordinates(ride.panel), st_coordinates(parks), 1))

## Distance to transit stop 
sfo <- read.csv('C:/Users/agarw/Documents/MUSA508/MUSA-508-Assignment-1/Data/final_sfo_lines.csv')
x <- vector(mode='list', length = 105)
y <- vector(mode='list', length = 105)
spli <- strsplit(sfo$Location, ",")
for (val in spli){
  x <- append(x, val[[1]])
  y <- append(y, val[[2]])
  
}
sfo_new <- 
  sfo %>%
  mutate(X = x[106:210],
         Y = y[106:210]) %>%
  dplyr::select(-Abbreviation, -Location, -Description)%>%
  st_as_sf(coords = c("X","Y"), crs = 4326, agr = "constant") %>%
  st_transform('ESRI:102241') %>%
  mutate(Legend = "Station")

ride.panel <-
  ride.panel %>% 
  mutate(
    station_dist = nn_function(st_coordinates(ride.panel), st_coordinates(sfo_new), 1))

## Tourist spots
tourist_spots <- st_read('Data/attraction_point.geojson') %>%
  st_transform('ESRI:102241')

ride.panel <-
  ride.panel %>% 
  mutate(
    tourist_dist = nn_function(st_coordinates(ride.panel), st_coordinates(tourist_spots), 1))

## Making the new features categorical 
ride.panel <-
  ride.panel %>%
  mutate(Park = case_when(
    park_dist > 0 & park_dist <= 500 ~ "Close",
    park_dist > 500 & park_dist <= 1000 ~ "Moderate",
    park_dist > 1000 ~ "Far"))

ride.panel <-
  ride.panel %>%
  mutate(Station = case_when(
    station_dist > 0 & station_dist <= 500 ~ "Close",
    station_dist > 500 & station_dist <= 1000 ~ "Moderate",
    station_dist > 1000 ~ "Far"))

ride.panel <-
  ride.panel %>%
  mutate(Tourist = case_when(
    tourist_dist > 0 & tourist_dist <= 500 ~ "Close",
    tourist_dist > 500 & tourist_dist <= 1000 ~ "Moderate",
    tourist_dist > 1000 ~ "Far"))
```

## 6.1 Linear Regressions

We split our data into a training and a test set. We create five linear models using the `lm` funtion. Sometimes, for data such as these, Poisson distributions, designed for modeling counts, might be appropriate. I'll spare you the effort - linear models work better with this particular data set. 

We create the models using our training data `ride.Train`. The first models include only temporal controls, but the later ones contain all of our lag information.

Notice that we are using a partition that is user specified and is time dependent. 

**Why are we splitting our data by time and not randomly??**

**Why are we using the end of the month to predict for the beginning of the month?**

Be aware that these models may run quite slowly depending on how large your data set is and how many variables you decide to include.

```{r train_test, cache = TRUE}
## Regression
ride.Train <- filter(ride.panel, week <= 12)
ride.Test <- filter(ride.panel, week > 12)

### 5 regressions
reg1 <- 
  lm(Trip_Count ~  hour(interval60) + dotw + Temperature,  data=ride.Train)

reg2 <- 
  lm(Trip_Count ~  start_station_name + dotw + Temperature,  data=ride.Train)

reg3 <- 
  lm(Trip_Count ~  start_station_name + hour(interval60) + dotw + Temperature + Precipitation, 
     data=ride.Train)

reg4 <- 
  lm(Trip_Count ~  start_station_name +  hour(interval60) + dotw + Temperature + Precipitation +
       lagHour + lag2Hours +lag3Hours + lag12Hours + lag1day, 
     data=ride.Train)

reg5 <- 
  lm(Trip_Count ~  start_station_name +  hour(interval60) + dotw + Temperature + Precipitation +
       lagHour + lag2Hours +lag3Hours + lag12Hours + lag1day + Park + Station + Tourist, 
     data=ride.Train)
```

## 6.2 Predict for test data

When your models have finished running, create a nested data frame of test data by week. Nested data is common in most other programming languages. For instance, the javascript object notation file format (aka JSON) is highly nested.

Nesting means that instead of merely having a "flat" file consisting of rows and columns, you have a matrix of other objects - imagine each cell in a matrix containing another matrix within it, or a list, or a list of lists. 

The `purrr` package is designed to `map` functions through nested data structures. This concept is important - think of `map` as visiting each dataframe in a nested data set and applies a function to it.

We create a function called `model_pred` which we can then `map` onto each data frame in our nested structure.

This function is called in the code below in a few ways, one way is like so: `map(.x = data, fit = name_of_your_regression, .f = model_pred)`. Here's the important bit - the argument `fit` takes the name of a regression you have created that you want to use to make predictions, and the `.f` argument takes a function, in this case `model_pred`, which we create in order to simply execute the `predict` function.

You don't need to manipulate anything else in this code block other than the `fit` argument.

```{r nest_data, cache = TRUE, warning = FALSE, message = FALSE}
## Predicting 
ride.Test.weekNest <- 
  ride.Test %>%
  nest(-week) 

model_pred <- function(dat, fit){
  pred <- predict(fit, newdata = dat)}

week_predictions <- 
  ride.Test.weekNest %>% 
  mutate(ATime_FE = map(.x = data, fit = reg1, .f = model_pred),
         BSpace_FE = map(.x = data, fit = reg2, .f = model_pred),
         CTime_Space_FE = map(.x = data, fit = reg3, .f = model_pred),
         DTime_Space_FE_timeLags = map(.x = data, fit = reg4, .f = model_pred),
         ETime_Space_FE_timeLags_Features = map(.x = data, fit = reg5, .f = model_pred)) %>% 
  gather(Regression, Prediction, -data, -week) %>%
  mutate(Observed = map(data, pull, Trip_Count),
         Absolute_Error = map2(Observed, Prediction,  ~ abs(.x - .y)),
         MAE = map_dbl(Absolute_Error, mean, na.rm = TRUE),
         sd_AE = map_dbl(Absolute_Error, sd, na.rm = TRUE))
```

## 6.3 Cross Validation

```{r cv, cache = TRUE, warning = FALSE, message = FALSE}
## Cross validation
library(caret)

bikenetsample <- sample_n(ride.panel, 10 0000)%>%
  na.omit()

fitControl <- trainControl(method = "cv", 
                           number = 100,
                           savePredictions = TRUE)

set.seed(1000)
# for k-folds CV

reg.cv <-  
  train(Trip_Count ~ start_station_name +  hour(interval60) + dotw + Temperature + Precipitation +
          lagHour + lag2Hours +lag3Hours + lag12Hours + lag1day, 
        data = bikenetsample,  
        method = "lm",  
        trControl = fitControl,  
        na.action = na.pass)

reg.cv
```

## 7 Accuracy

### 7.1 General Error Metrics 
```{r err_1, cache = TRUE, warning = FALSE, message = FALSE}
week_predictions <- 
  ride.Test.weekNest %>% 
  mutate(ATime_FE = map(.x = data, fit = reg1, .f = model_pred),
         BSpace_FE = map(.x = data, fit = reg2, .f = model_pred),
         CTime_Space_FE = map(.x = data, fit = reg3, .f = model_pred),
         DTime_Space_FE_timeLags = map(.x = data, fit = reg4, .f = model_pred),
         ETime_Space_FE_timeLags_Features = map(.x = data, fit = reg5, .f = model_pred)) %>% 
  gather(Regression, Prediction, -data, -week) %>%
  mutate(Observed = map(data, pull, Trip_Count),
         Absolute_Error = map2(Observed, Prediction,  ~ abs(.x - .y)),
         MAE = map_dbl(Absolute_Error, mean, na.rm = TRUE),
         sd_AE = map_dbl(Absolute_Error, sd, na.rm = TRUE))

## Bar plot
week_predictions %>%
  dplyr::select(week, Regression, MAE) %>%
  gather(Variable, MAE, -Regression, -week) %>%
  ggplot(aes(week, MAE)) + 
  geom_bar(aes(fill = Regression), position = "dodge", stat="identity") +
  scale_fill_manual(values = palette5) +
  labs(title = "Mean Absolute Errors by model specification and week") +
  plotTheme()

## Graph
week_predictions %>% 
  mutate(interval60 = map(data, pull, interval60),
         start_station_id = map(data, pull, start_station_id)) %>%
  dplyr::select(interval60, start_station_id, Observed, Prediction, Regression) %>%
  unnest() %>%
  gather(Variable, Value, -Regression, -interval60, -start_station_id) %>%
  group_by(Regression, Variable, interval60) %>%
  summarize(Value = sum(Value)) %>%
  ggplot(aes(interval60, Value, colour=Variable)) + 
  geom_line(size = 1.1) + 
  facet_wrap(~Regression, ncol=1) +
  labs(title = "Predicted/Observed bike share time series", subtitle = "SF + Alameda",  x = "Hour", y= "Station Trips") +
  plotTheme()
```

### 7.2 Specific Error Metrics
```{r err_2, cache = TRUE, warning = FALSE, message = FALSE}
## MAE by station
week_predictions %>% 
  mutate(interval60 = map(data, pull, interval60),
         start_station_id = map(data, pull, start_station_id), 
         start_station_latitude = map(data, pull, start_station_latitude), 
         start_station_longitude = map(data, pull, start_station_longitude)) %>%
  select(interval60, start_station_id, start_station_longitude, start_station_latitude, Observed, Prediction, Regression) %>%
  unnest() %>%
  filter(Regression == "DTime_Space_FE_timeLags") %>%
  group_by(start_station_id, start_station_longitude, start_station_latitude) %>%
  summarize(MAE = mean(abs(Observed-Prediction), na.rm = TRUE))%>%
  ggplot(.)+
  geom_sf(data = sfTracts %>%
            st_transform(crs=4326), colour = '#efefef')+
  geom_point(aes(x = start_station_longitude, y = start_station_latitude, color = MAE), 
             fill = "transparent", alpha = 0.4)+
  scale_colour_viridis(direction = -1,
                       discrete = FALSE, option = "D")+
  ylim(min(sfBike_census$start_station_latitude), max(sfBike_census$start_station_latitude))+
  xlim(min(sfBike_census$start_station_longitude), max(sfBike_census$start_station_longitude))+
  labs(title="Mean Abs Error, Test Set, Model 4")+
  mapTheme()
```


```{r err_3, cache = TRUE, warning = FALSE, message = FALSE}
## Scatterplot
week_predictions %>% 
  mutate(interval60 = map(data, pull, interval60),
         start_station_id = map(data, pull, start_station_id), 
         start_station_latitude = map(data, pull, start_station_latitude), 
         start_station_longitude = map(data, pull, start_station_longitude),
         dotw = map(data, pull, dotw)) %>%
  select(interval60, start_station_id, start_station_longitude, 
         start_station_latitude, Observed, Prediction, Regression,
         dotw) %>%
  unnest() %>%
  filter(Regression == "DTime_Space_FE_timeLags")%>%
  mutate(weekend = ifelse(dotw %in% c("Sun", "Sat"), "Weekend", "Weekday"),
         time_of_day = case_when(hour(interval60) < 7 | hour(interval60) > 18 ~ "Overnight",
                                 hour(interval60) >= 7 & hour(interval60) < 10 ~ "AM Rush",
                                 hour(interval60) >= 10 & hour(interval60) < 15 ~ "Mid-Day",
                                 hour(interval60) >= 15 & hour(interval60) <= 18 ~ "PM Rush"))%>%
  ggplot()+
  geom_point(aes(x= Observed, y = Prediction))+
  geom_smooth(aes(x= Observed, y= Prediction), method = "lm", se = FALSE, color = "red")+
  geom_abline(slope = 1, intercept = 0)+
  facet_grid(time_of_day~weekend)+
  labs(title="Observed vs Predicted",
       x="Observed trips", 
       y="Predicted trips")+
  plotTheme()
```


```{r err_4, cache = TRUE, warning = FALSE, message = FALSE}
##MAE by hour
week_predictions %>% 
  mutate(interval60 = map(data, pull, interval60),
         start_station_id = map(data, pull, start_station_id), 
         start_station_latitude = map(data, pull, start_station_latitude), 
         start_station_longitude = map(data, pull, start_station_longitude),
         dotw = map(data, pull, dotw) ) %>%
  select(interval60, start_station_id, start_station_longitude, 
         start_station_latitude, Observed, Prediction, Regression,
         dotw) %>%
  unnest() %>%
  filter(Regression == "DTime_Space_FE_timeLags")%>%
  mutate(weekend = ifelse(dotw %in% c("Sun", "Sat"), "Weekend", "Weekday"),
         time_of_day = case_when(hour(interval60) < 7 | hour(interval60) > 18 ~ "Overnight",
                                 hour(interval60) >= 7 & hour(interval60) < 10 ~ "AM Rush",
                                 hour(interval60) >= 10 & hour(interval60) < 15 ~ "Mid-Day",
                                 hour(interval60) >= 15 & hour(interval60) <= 18 ~ "PM Rush")) %>%
  group_by(start_station_id, weekend, time_of_day, start_station_longitude, start_station_latitude) %>%
  summarize(MAE = mean(abs(Observed-Prediction), na.rm = TRUE))%>%
  ggplot(.)+
  geom_sf(data = sfTracts %>%
            st_transform(crs=4326), colour = '#efefef')+
  geom_point(aes(x = start_station_longitude, y = start_station_latitude, color = MAE), 
             fill = "transparent", size = 0.5, alpha = 1.5)+
  scale_colour_viridis(direction = -1,
                       discrete = FALSE, option = "D")+
  ylim(min(sfBike_census$start_station_latitude), max(sfBike_census$start_station_latitude))+
  xlim(min(sfBike_census$start_station_longitude), max(sfBike_census$start_station_longitude))+
  facet_grid(weekend~time_of_day)+
  labs(title="Mean Absolute Errors, Test Set")+
  mapTheme()
```


```{r err_5, cache = TRUE, warning = FALSE, message = FALSE}
## Error and census
week_predictions %>% 
  mutate(interval60 = map(data, pull, interval60),
         start_station_id = map(data, pull, start_station_id), 
         start_station_latitude = map(data, pull, start_station_latitude), 
         start_station_longitude = map(data, pull, start_station_longitude),
         dotw = map(data, pull, dotw),
         Percent_Taking_Public_Trans = map(data, pull, Percent_Taking_Public_Trans),
         Med_Inc = map(data, pull, Med_Inc),
         Percent_White = map(data, pull, Percent_White)) %>%
  select(interval60, start_station_id, start_station_longitude, 
         start_station_latitude, Observed, Prediction, Regression,
         dotw, Percent_Taking_Public_Trans, Med_Inc, Percent_White) %>%
  unnest() %>%
  filter(Regression == "DTime_Space_FE_timeLags")%>%
  mutate(weekend = ifelse(dotw %in% c("Sun", "Sat"), "Weekend", "Weekday"),
         time_of_day = case_when(hour(interval60) < 7 | hour(interval60) > 18 ~ "Overnight",
                                 hour(interval60) >= 7 & hour(interval60) < 10 ~ "AM Rush",
                                 hour(interval60) >= 10 & hour(interval60) < 15 ~ "Mid-Day",
                                 hour(interval60) >= 15 & hour(interval60) <= 18 ~ "PM Rush")) %>%
  filter(time_of_day == "AM Rush") %>%
  group_by(start_station_id, Percent_Taking_Public_Trans, Med_Inc, Percent_White) %>%
  summarize(MAE = mean(abs(Observed-Prediction), na.rm = TRUE))%>%
  gather(-start_station_id, -MAE, key = "variable", value = "value")%>%
  ggplot(.)+
  #geom_sf(data = sfCensus, color = "grey", fill = "transparent")+
  geom_point(aes(x = value, y = MAE), alpha = 0.4)+
  geom_smooth(aes(x = value, y = MAE), method = "lm", se= FALSE)+
  facet_wrap(~variable, scales = "free")+
  labs(title="Errors as a function of socio-economic variables",
       y="Mean Absolute Error (Trips)")+
  plotTheme()
```


```{r animation, cache = TRUE, warning = FALSE, message = FALSE}
## Animation

library(gganimate)
library(gifski)

week11 <-
  filter(sfBike_census , week == 11)

week11.panel <-
  expand.grid(
    interval15 = unique(week11$interval15),
    Pickup.Census.Tract = unique(sfBike_census$start_station_id))

ride.animation.data <-
  mutate(week11, Trip_Counter = 1) %>%
  select(interval15, start_station_id, start_station_longitude, start_station_latitude, Trip_Counter) %>%
  group_by(interval15, start_station_id, start_station_longitude, start_station_latitude) %>%
  summarize(Trip_Count = sum(Trip_Counter, na.rm=T)) %>% 
  ungroup() %>% 
  mutate(Trips = case_when(Trip_Count == 0 ~ "0 trips",
                           Trip_Count > 0 & Trip_Count <= 2 ~ "0-2 trips",
                           Trip_Count > 2 & Trip_Count <= 5 ~ "2-5 trips",
                           Trip_Count > 5 & Trip_Count <= 10 ~ "5-10 trips",
                           Trip_Count > 10 & Trip_Count <= 15 ~ "10-15 trips",
                           Trip_Count > 15 ~ "15+ trips")) %>%
  mutate(Trips  = fct_relevel(Trips, "0 trips","0-2 trips","2-5 trips",
                              "5-10 trips","10-15 trips","15+ trips"))

rideshare_animation <-
  ggplot()+
  geom_sf(data = sfTracts %>%
            st_transform(crs=4326), colour = '#efefef')+
  geom_point(data = ride.animation.data, 
             aes(x = start_station_longitude, y = start_station_latitude, fill = Trips), size = 0.5, alpha = 1.5) +
  scale_colour_manual(values = palette5) +
  labs(title = "Rideshare pickups for one week in March 2018",
       subtitle = "15 minute intervals: {current_frame}") +
  transition_manual(interval15) +
  mapTheme()

animate(rideshare_animation, duration=20, renderer = gifski_renderer())
```