---
title: "Space-Time Prediction of Bike Share Demand"
author: "Palak Agarwal"
date: "November 13, 2020"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: "hide"
    code_download: true
---

## 1 Introduction

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The San Francisco Municipal Transportation Agency (SFMTA) is a department of the City and County of San Francisco responsible for the management of all ground transportation in the city. The SFMTA has oversight over the Municipal Railway (Muni) public transit, as well as bicycling, paratransit, parking, traffic, walking, and taxis. In 2013 they partnered with public agencies to plan and implement the original Bay Area Bike Share pilot project and are now leading San Francisco’s efforts to work with the private sector partner to expand the system to over ten times its pilot size. They patnered with Lyft to expand the regional system which has since been re-branded with all-new equipment as Bay Wheels. In 2017, as independent stationless bikeshare emerged as a big new trend, they were among the first U.S. cities to create a regulatory and permitting framework to address this fast-moving phenomenon and insure that bikeshare in all its forms is safe, orderly and equitable for all San Franciscans.

One of the big operational challenges of bike share systems is "re-balancing" - getting bikes to stations that are anticipated to have demand but lack bikes. Figuring out how to do this is one of the keys to operating a successful system. With the introduction of the stationless bikeshare system the problem of re-balancing can be eliminated by a certain percentage. But it creates new problems as the bikes are further spread out and its harder to make sure that the bike is picked up. For the assignment, the data used is only from docked stations and does not include the stationless bikes.

For making the bike-share system more efficient i.e. to re-balance it many strategies can be introduced. I think the most effect strategy would be to use a reward system. If we know the capacity of a bike station and knew that the user was going to that station we could incentivize them to drop off the bike to another bike station or we could also incentivize them to use a bike which has not been touched for more than a two weeks. This will allow for an efficient demand and supply system.

Today's trend is similar to the trend tomorrow and this week's trend will be similar to next week trends. Hence, we will be able to predict how many bikes and from which station to high accuracy as you will see in the later in the model and assignment. If we know how many bikes are going to be in use, it is easier to use that data to manipulate which bikes are being used and which are not. For example, the Financial District of downtown SF is very busy and has many stations in the same block. If for some reason one station is more active than the other, we can use rewards to make people pick and drop bikes from different stations. 

Lyft using its platform and app to show the users where the bikes are available so that can be used to make users pick up bikes that have not been touched for a while and also have them drop it off to locations where the demand is high and there is a shortage of bikes. SFMTA already has concessions for different individuals and communities which the bike share system can tap into as well. 

## 2.0 Setup
### 2.1 Libraries and themes
Let's load relevant libraries and some graphic themes.
```{r setup, cache=TRUE, message=FALSE}
## Setup
library(tidyverse)
library(sf)
library(lubridate)
library(tigris)
library(tidycensus)
library(viridis)
library(riem)
library(gridExtra)
library(knitr)
library(kableExtra)
library(RSocrata)

mapTheme <- function(base_size = 12) {
  theme(
    text = element_text( color = "black"),
    plot.title = element_text(size = 24,colour = "black"),
    plot.subtitle=element_text(face="italic"),
    plot.caption=element_text(hjust=0),
    axis.ticks = element_blank(),
    panel.background = element_blank(),axis.title = element_blank(),
    axis.text = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill=NA, size=2)
  )
}

plotTheme <- function(base_size = 12) {
  theme(
    text = element_text( color = "black"),
    plot.title = element_text(size = 24,colour = "black"),
    plot.subtitle = element_text(face="italic"),
    plot.caption = element_text(hjust=0),
    axis.ticks = element_blank(),
    panel.background = element_blank(),
    panel.grid.major = element_line("grey60", size = 0.1),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill=NA, size=2),
    strip.background = element_rect(fill = "grey60", color = "white"),
    strip.text = element_text(size=12),
    axis.title = element_text(size=12),
    axis.text = element_text(size=10),
    plot.background = element_blank(),
    legend.background = element_blank(),
    legend.title = element_text(colour = "black", face = "italic"),
    legend.text = element_text(colour = "black", face = "italic"),
    strip.text.x = element_text(size = 14)
  )
}

# Load Quantile break functions

qBr <- function(df, variable, rnd) {
  if (missing(rnd)) {
    as.character(quantile(round(df[[variable]],0),
                          c(.01,.2,.4,.6,.8), na.rm=T))
  } else if (rnd == FALSE | rnd == F) {
    as.character(formatC(quantile(df[[variable]]), digits = 3),
                 c(.01,.2,.4,.6,.8), na.rm=T)
  }
}

q5 <- function(variable) {as.factor(ntile(variable, 5))}

palette5 <- c("#eff3ff","#bdd7e7","#6baed6","#3182bd","#08519c")
palette4 <- c("#D2FBD4","#92BCAB","#527D82","#123F5A")
palette2 <- c("#6baed6","#08519c")
```

### 2.2 Data Wrangling
Reading in the bike data.
```{r load_data, message = FALSE}
## Read data
sf_bike1 <- read.csv('Data/201803-fordgobike.csv')
sf_bike2 <- read.csv('Data/201804-fordgobike.csv')

sf_bike <- rbind(sf_bike1, sf_bike2)
glimpse(sf_bike)

sfCounty <-
  st_read("https://data.sfgov.org/api/geospatial/p5b7-5n3h?method=export&format=GeoJSON") %>% 
  st_union() %>%
  st_transform('ESRI:102241')

neighborhoods <- 
  st_read("https://data.sfgov.org/api/geospatial/pty2-tcw4?method=export&format=GeoJSON") %>%
  st_transform('ESRI:102241') %>%
  dplyr::select(-link)
```
After that we use the data to create bins of 15 and 60 minute intervals by rounding.
```{r time_bins, cache = TRUE}
sfBike <- sf_bike %>%
  mutate(interval60 = floor_date(ymd_hms(start_time), unit = "hour"),
         interval15 = floor_date(ymd_hms(start_time), unit = "15 mins"),
         week = week(interval60),
         dotw = wday(interval60, label=TRUE)) %>%
  filter(week < 15)
```

### 2.3 Census Data
```{r census, message = FALSE}
# Install Census API Key
census_api_key("aea3dee2d96acb5101e94f3dcfa1b575f73d093a", overwrite = TRUE)

sfCensus <- 
  get_acs(geography = "tract", 
          variables = c("B01003_001", "B19013_001", 
                        "B02001_002", "B08013_001",
                        "B08012_001", "B08301_001", 
                        "B08301_010", "B01002_001"), 
          year = 2018, 
          state = 06, 
          geometry = TRUE, 
          county=c(001,075,081,013,085),
          output = "wide") %>%
  rename(Total_Pop =  B01003_001E,
         Med_Inc = B19013_001E,
         Med_Age = B01002_001E,
         White_Pop = B02001_002E,
         Travel_Time = B08013_001E,
         Num_Commuters = B08012_001E,
         Means_of_Transport = B08301_001E,
         Total_Public_Trans = B08301_010E) %>%
  select(Total_Pop, Med_Inc, White_Pop, Travel_Time,
         Means_of_Transport, Total_Public_Trans,
         Med_Age,
         GEOID, geometry) %>%
  mutate(Percent_White = White_Pop / Total_Pop,
         Mean_Commute_Time = Travel_Time / Total_Public_Trans,
         Percent_Taking_Public_Trans = Total_Public_Trans / Means_of_Transport) %>%
  st_transform('ESRI:102241')

sfTracts <- 
  sfCensus %>%
  as.data.frame() %>%
  distinct(GEOID, .keep_all = TRUE) %>%
  select(GEOID, geometry) %>% 
  st_sf
```

## 2.4 Connect census data to the the Bikeshare data
Now lets add the spatial information to our bikeshare data as origin and destination data, first joining the origin station, then the destination station to our census data.
```{r add_census_tracts, cache = TRUE, message = FALSE, warning = FALSE}
## Adding census data to bike data
sfBike_census <- st_join(sfBike %>% 
                        filter(is.na(start_station_longitude) == FALSE &
                                 is.na(start_station_latitude) == FALSE &
                                 is.na(end_station_latitude) == FALSE &
                                 is.na(end_station_longitude) == FALSE) %>%
                        st_as_sf(., coords = c("start_station_longitude", "start_station_latitude"), crs = 4326),
                      sfTracts %>%
                        st_transform(crs=4326),
                      join=st_intersects,
                          left = TRUE) %>%
  rename(Origin.Tract = GEOID) %>%
  mutate(start_station_longitude = unlist(map(geometry, 1)),
         start_station_latitude = unlist(map(geometry, 2)))%>%
  as.data.frame() %>%
  select(-geometry)%>%
  st_as_sf(., coords = c("end_station_longitude", "end_station_latitude"), crs = 4326) %>%
  st_join(., sfTracts %>%
            st_transform(crs=4326),
          join=st_intersects,
          left = TRUE) %>%
  rename(Destination.Tract = GEOID)  %>%
  mutate(end_station_longitude = unlist(map(geometry, 1)),
         end_station_latitude = unlist(map(geometry, 2)))%>%
  as.data.frame() %>%
  select(-geometry)
```